The socialization of meritocracy and market justice preferences at school

### 2024-10-02 01-introduction.qmd
Update revised version MPI Submit

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus on justifying inequalities in health, education and pensions, which are traditional social policy areas where access to better services could be justified based on payment capacity, referred to as market justice [@lane_market_1986; @lindh_bringing_2023]. Market justice preferences refers to a belief in a market-based distributive principle, where goods and rewards are allocated according to individual merit, effort and contributions in a free-market economy (Lindh, 2015).

Most of the research on meritocracy and justice preferences has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Studying beliefs about meritocracy and market justice preferences at school age is crucial because these early beliefs can shape individuals' attitudes towards inequality and social justice later in life. As Castillo (2019) argues, meritocratic ideals "serve as a principle that legitimizes the distribution of goods and rewards based on individual talent and effort," making it essential to understand how these beliefs develop during adolescence. Furthermore, early exposure to these beliefs may influence broader perceptions of fairness, as Reynolds (2014) highlights, "beliefs about the stratification system influence people's judgments about the fairness of inequality," suggesting that such attitudes formed at a young age can have lasting impacts. Additionally, García-Sierra (2023) emphasizes that the effects of holding meritocratic beliefs are not uniform, as these ideologies may reinforce socioeconomic disparities depending on students' social backgrounds.

>>>>

In the present paper, we address the role of the perception of meritocracy on the justification of social inequalities by eighth-grade students in Chile, a country characterized by a highly stratified educational system. In particular, we focus on justifying inequalities in health, education and pensions, which are traditional social policy areas where access to better services could be justified based on payment capacity, referred to as market justice [@lane_market_1986; @lindh_bringing_2023]. Market justice preferences refers to a belief in a market-based distributive principle, where goods and rewards are allocated according to individual merit, effort and contributions in a free-market economy [@lindh_public_2015].

Most of the research on meritocracy and justice preferences has only considered adults, leaving aside the study of how beliefs in this field develop at student age as well as the impact of the school context and the family as the main socialization agencies [@batruch_belief_2022]. Studying beliefs about meritocracy and market justice preferences at school age is crucial because these early beliefs can shape individuals' attitudes towards inequality and social justice later in life. As Castillo et al. [-@castillo_meritocracia_2019] argues, meritocratic ideals serve as a principle that legitimizes the distribution of goods and rewards based on individual talent and effort, making it essential to understand how these beliefs develop during adolescence. Furthermore, early exposure to these beliefs may influence broader perceptions of fairness, as Reynolds and Xian [-@reynolds_perceptions_2014, p.134] highlights, beliefs about the stratification system influence people's judgments about the fairness of inequality, suggesting that such attitudes formed at a young age can have lasting impacts. Additionally, García-Sierra [-@garcia-sierra_dark_2023] emphasizes that the effects of holding meritocratic beliefs are not uniform, as these ideologies may reinforce socioeconomic disparities depending on students' social backgrounds.

### 2024-10-02 cover_letter.qmd
Update revised version MPI Submit

---
title: "The Socialization of Meritocracy and Market Justice Preferences at School"
subtitle: "Cover Letter"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
  hikmah-manuscript-pdf: 
    indent: false
    keep-tex: true
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.

@@

Mauricio Salgado, School of Sociology, Universidad Andrés Bello.  
ORCID: 0000-0002-8108-6154

>>>>

---
title: "The Socialization of Meritocracy and Market Justice Preferences at School"
subtitle: "Cover Letter [Revised Version]"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
  hikmah-manuscript-pdf: 
    indent: false
    keep-tex: true
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

Previous research has shown that schools often justify student performance differences using meritocratic ideals. One potential consequence of such ideals is the legitimization of outcome inequalities across various spheres, including those traditionally associated with equality and redistribution. In this study, we argue that the promotion of meritocratic values during school-age can shape students' beliefs about meritocracy and influence their views on market-based access to health, pensions, and education. Using data from the 2017 National Study of Civic Education in Chile, which includes 5,047 8th-grade students from 231 schools, we estimated a series of multilevel models (lme4 library, R version 4.1.3) to test our hypotheses. Our findings show that a significant proportion of Chilean students agree with market justice principles—more so than adults. Most students endorse meritocratic views, particularly the notion that effort should be rewarded, which strongly correlates with market justice preferences: students who believe in meritocracy are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with higher academic achievement. Furthermore, the conditional influence of meritocratic beliefs diminishes in schools with higher socioeconomic status and performance levels. These results suggest that the association between meritocratic beliefs and market justice preferences is already established at school age and is shaped by the school environment.

@@

Mauricio Salgado, School of Sociology, Universidad Andrés Bello and Centro de Estudios Públicos (CEP).
ORCID: 0000-0002-8108-6154

### 2024-10-02 cover_letter.tex
Update revised version MPI Submit

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\vskip 0.25em \large #1 \par}}{}{}
}
\makeatother
\subtitle{Cover Letter}

@@

\setstretch{1.15}
Previous research has shown that schools justify student performance
differences based on meritocratic ideals. However, the link between
school meritocracy and students' justification of inequalities has
received little attention. This paper examines how school and family
socialization influence students' beliefs about meritocracy and
market-based justifications for access to health, pensions, and
education in Chile. Using data from the First Study of Civic Education
in Chile, we employ a multilevel modeling strategy to analyze 5,047
students and parents nested within 231 schools. Our findings reveal that
a significant proportion of Chilean students agree with market justice
principles, at a higher rate than adults. Most students endorse
meritocratic views, particularly the reward of effort. Multilevel
estimations show strong associations between meritocratic beliefs and
market justice preferences: students who believe in the reward of effort
and talent are more likely to justify inequalities based on financial
capacity. At the school level, market justice preferences are higher in
high-status schools but lower in schools with better achievement.
Additionally, the conditional influence of meritocracy beliefs weakens
in schools with higher socioeconomic status and performance. These
results suggest that meritocratic beliefs and market justice preferences
begin early and are influenced by the school environment.

@@

Mauricio Salgado, School of Sociology, Universidad Andrés Bello.\\
ORCID: 0000-0002-8108-6154

>>>>

\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\vskip 0.25em \large #1 \par}}{}{}
}
\makeatother
\subtitle{Cover Letter {[}Revised Version{]}}

@@

\setstretch{1.15}
Previous research has shown that schools often justify student
performance differences using meritocratic ideals. One potential
consequence of such ideals is the legitimization of outcome inequalities
across various spheres, including those traditionally associated with
equality and redistribution. In this study, we argue that the promotion
of meritocratic values during school-age can shape students' beliefs
about meritocracy and influence their views on market-based access to
health, pensions, and education. Using data from the 2017 National Study
of Civic Education in Chile, which includes 5,047 8th-grade students
from 231 schools, we estimated a series of multilevel models (lme4
library, R version 4.1.3) to test our hypotheses. Our findings show that
a significant proportion of Chilean students agree with market justice
principles---more so than adults. Most students endorse meritocratic
views, particularly the notion that effort should be rewarded, which
strongly correlates with market justice preferences: students who
believe in meritocracy are more likely to justify inequalities based on
financial capacity. At the school level, market justice preferences are
higher in high-status schools but lower in schools with higher academic
achievement. Furthermore, the conditional influence of meritocratic
beliefs diminishes in schools with higher socioeconomic status and
performance levels. These results suggest that the association between
meritocratic beliefs and market justice preferences is already
established at school age and is shaped by the school environment.

@@

Mauricio Salgado, School of Sociology, Universidad Andrés Bello and
Centro de Estudios Públicos (CEP). ORCID: 0000-0002-8108-6154

### 2024-10-02 paper-blinded.qmd
Update revised version MPI Submit

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
  hikmah-manuscript-pdf: 
    indent: false
    keep-tex: true
abstract: |
  Previous research has shown that schools often justify student performance differences using meritocratic ideals. One potential consequence of such ideals is the legitimization of outcome inequalities across various spheres, including those traditionally associated with equality and redistribution. In this study, we argue that the promotion of meritocratic values during school-age can shape students' beliefs about meritocracy and influence their views on market-based access to health, pensions, and education. Using data from the 2017 National Study of Civic Education in Chile, which includes 5,047 8th-grade students from 231 schools, we estimated a series of multilevel models (lme4 library, R version 4.1.3) to test our hypotheses. Our findings show that a significant proportion of Chilean students agree with market justice principles—more so than adults. Most students endorse meritocratic views, particularly the notion that effort should be rewarded, which strongly correlates with market justice preferences: students who believe in meritocracy are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with higher academic achievement. Furthermore, the conditional influence of meritocratic beliefs diminishes in schools with higher socioeconomic status and performance levels. These results suggest that the association between meritocratic beliefs and market justice preferences is already established at school age and is shaped by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---
{{< pagebreak >}}

>>>>

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
    keep-tex: true
abstract: |
  Previous research has shown that schools often justify student performance differences using meritocratic ideals. One potential consequence of such ideals is the legitimization of outcome inequalities across various spheres, including those traditionally associated with equality and redistribution. In this study, we argue that the promotion of meritocratic values during school-age can shape students' beliefs about meritocracy and influence their views on market-based access to health, pensions, and education. Using data from the 2017 National Study of Civic Education in Chile, which includes 5,047 8th-grade students from 231 schools, we estimated a series of multilevel models (lme4 library, R version 4.1.3) to test our hypotheses. Our findings show that a significant proportion of Chilean students agree with market justice principles—more so than adults. Most students endorse meritocratic views, particularly the notion that effort should be rewarded, which strongly correlates with market justice preferences: students who believe in meritocracy are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with higher academic achievement. Furthermore, the conditional influence of meritocratic beliefs diminishes in schools with higher socioeconomic status and performance levels. These results suggest that the association between meritocratic beliefs and market justice preferences is already established at school age and is shaped by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
number-sections: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---
{{< pagebreak >}}
